### 2024-01-12 pnrs.tex
correcting some mistakes

When multiple indeterminate are involved, a bit of care is needed, and
this is the technical content of \cref{diff-one-monomial:lem,lem:delta-cool}.

\begin{definition}
    Let $\vec{X}$ be a tuple of indeterminate, let $Y$
    be a fresh indeterminate,
    and 
    let $P \in \Rel[\vec{X},Y]$ be a polynomial.
    For every \kl{maximal monomial}
    $\alpha Y^p M$ of $P$ such that $M \in \Rel[\vec{X}]$, and every $K \in \Nat$,
    we define
    \begin{equation*}
        \Diff{K}{\alpha Y^p M}{P} \defined 
        P - \alpha (Y - K)^p M \quad .
    \end{equation*}
\end{definition}

\begin{lemma}
    \label{diff-one-monomial:lem}
    Let $\vec{X}$ be a tuple of indeterminate, $Y$ be a fresh indeterminate,
    and $P = Y^p M$ where $M$ is a \kl{monomial} over $\vec{X}$.
    Then, for all $K \geq 0$,
    for all $\nu \colon \vec{X}Y \topartial \Nat$
    such that $\nu(Y) > K$ or $\nu(Y)$ is undefined,
    $\restr{\Diff{K}{Y^pM}{Y^pM}}{\nu}$
    has \kl{maximal monomials} with
    coefficients at least $K$, or is the constant polynomial equal to zero.
\end{lemma}
\begin{proof}
    Let us write $\Delta_K$ as a shorthand for
    $\Diff{K}{Y^p M}{Y^p M}$.
    Let us first remark that
    there exists $Q \in \Rel[Y]$ of degree at most $p-2$
    such that:
    \begin{equation*}
        \Delta_K
        = KY^{p-1} M + Q M
        \quad .
    \end{equation*}
    In particular, there is a unique \kl{maximal monomial} for $\Delta_{K}$,
    which is precisely $K Y^{p-1} M$, because all monomials
    in $Q M$ \kl{strictly divide} $Y^{p-1} M$.

    Let $\nu \colon \vec{X}Y \topartial \Nat$ be a partial valuation such that
    $\restr{\Delta_K}{\nu} \neq 0$. If $\nu(Y)$ is undefined, then the
    \kl{maximal monomial} of $\restr{\Delta_K}{\nu}$ is exactly $K Y^{p-1}
    \restr{M}{\nu}$, and $\restr{M}{\nu} \neq 0$. Therefore, the unique
    \kl{maximal monomial} of $\restr{\Delta_K}{\nu}$ has a \kl{positive
    coefficient} that is at least $K$.

    Otherwise, if $\nu(Y) > K$, then $\nu(Y) = K + l$ for some $l \in \Nat$.
    Let us write $\mu$ for the restriction of $\nu$ to indeterminates of $\vec{X}$.
    We can compute 
    \begin{align*}
        \restr{\Delta_K}{\nu}
        &= (K + l)^p \restr{M}{\mu} - l^p \restr{M}{\mu}\\
        &= ((K + l)^p - l^p) \restr{M}{\mu} \\
        &= (K + l - l)\left(\sum_{i = 0}^{p-1} (K+l)^i l^{p-i}\right) 
        \restr{M}{\mu} \\
        &= K \alpha \restr{M}{\mu} 
        \quad .
    \end{align*}
    Which proves that the \kl{maximal monomial} of
    $\restr{\Delta_K}{\nu}$ has a coefficient
    that is at least $K$.
\end{proof}

We can refine the above lemma to use it in the case where
we do not have a single
monomial.

\begin{lemma}
    \label{lem:delta-cool}
    Let $\vec{X}$ be a tuple of indeterminate, and $Y$ be a fresh indeterminate.
    Let $P \in \Rel[\vec{X}Y]$ be in $\CorrectPoly$, and $\alpha Y^p M$ be a
    \kl{maximal monomial} of $P$, with $\alpha > 0$.
    Then, there exists a computable $K \geq 0$
    such that
    for all $\nu \colon \vec{X}Y \topartial \Nat$
    such that $\nu(Y) > K$ or $\nu(Y)$ is undefined,
    the \kl{maximal monomials} of 
    $\restr{\Diff{K}{Y^pM}{P}}{\nu}$
    have \kl{positive coefficients}.
\end{lemma}
\begin{proof}
    Let $N$ be greater than the sum of the absolute values of all coefficients
    appearing in $P$, and let $K \defined N + 1$.
    Let us write $\Delta_K$
    as a shorthand for
    $\Diff{K}{\alpha Y^p M}{P}$, and decompose $\Delta_K$ as follows:
    \begin{equation*}
        \Delta_K
        = \underbrace{(P - \alpha Y^p M)}_{Q_1}
        + \underbrace{\alpha (Y^p M - (Y - K)^p M)}_{Q_2}
        \quad .
    \end{equation*}

    Applying \cref{diff-one-monomial:lem}, we know that for every partial
    evaluation $\nu \colon \vec{X}Y \topartial \Nat$ such that $\nu(Y) > K$ or
    $\nu(Y)$ is undefined, the \kl{maximal monomials} of $\restr{Q_2}{\nu}$
    have \kl{positive coefficients}, and that these coefficients are either $0$
    (the polynomial is constant equal to zero) or at least $K$.

    To conclude, it is just a matter of describing how \kl{maximal monomials}
    of $\restr{\Delta_K}{\nu}$ can appear. Let us proceed by case analysis. If
    $\restr{Q_2}{\nu} = 0$, then, $\restr{\alpha Y^p M}{\nu} = 0$, and
    $\restr{\Delta_K}{\nu} = \restr{P}{\nu}$, in which case we conclude because
    $P \in \CorrectPoly$. Otherwise, we know that $\restr{Q_2}{\nu} \neq 0$,
    and that its (unique) \kl{maximal monomial} has a \kl{positive coefficient}
    that is at least $K$.
    Let us now consider a \kl{maximal monomial} $M_0$
    of $\restr{P}{\nu}$. This \kl{monomial} is obtained
    by summing \kl{monomials} from $\restr{Q_1}{\nu}$ and $\restr{Q_2}{\nu}$.
    Note that, by construction

    Because $\alpha Y^p M$ is a \kl{maximal monomial} of
    $P$, and since $K$ is greater than the sum of
    the absolute values of all coefficients appearing in $P$,
    we conclude that

    If a \kl{maximal monomial} of $\restr{\Delta_K}{\nu}$
    is obtained
    by summing a of $\restr{Q_1}{\nu}$
    by summing an element of the left hand side with an element
    of the right hand side, it needs to be a \kl{maximal monomial}
    of the right hand side, and the large value of $K$ guarantees
    that the sum of coefficients is positive.
    If it only appears on the right hand side, it is even better.
    If it only appears on the left hand side, then
    in particular it means that the monomial does not
    divide $\restr{Y^p M}{\nu}$, hence has the same value
    in $\restr{\Delta_K}{\nu}$ or in $\restr{P}{\nu}$,
    which is guaranteed to be positive because $P \in \CorrectPoly$.

    In particular, we conclude from this that
    $\Delta_Y \geq 0$, on all tuples such that $Y > K$.
\end{proof}

Note that we cannot conclude in the above lemma that the result is in
$\CorrectPoly$, because we have a restriction on the range where the variable
$Y$ can be taken. The upcoming \cref{lem:correct-to-n-rat} is proven using an
induction on the number $d$ of indeterminate, and the degrees of the
\kl{maximal monomials}. To make the induction formal, let us define
$\Monomials[\vec{X}]$ as the (infinite) set of all possible \kl{monomials} over
$\vec{X}$, ordered by \kl{divisibility}. Remark that this set is isomorphic to
$\Nat^{\vec{X}}$ with the product ordering, which is a
\kl{well-quasi-ordering}.

@@

@@

    \textbf{Induction:}
    Assume that $\alpha Y^p M \in \MaximalMonomials(P)$
    for some indeterminate $Y$, $\alpha \neq 0$, and $p \geq 1$.
    Because $P \in \CorrectPoly$, we know that $\alpha > 0$.
    Using \cref{lem:delta-cool},
    we know that there exists a computable $K$ such that
    the polynomial
    $\Diff{K} \defined P - \alpha (Y - K)^p M$
    satisfies that for all partial valuations $\nu \colon Y\vec{X} \topartial \Nat$
    such that $\nu(Y) > K$ or $\nu(Y)$ is undefined,
    the \kl{maximal monomials} of $\Diff{K}$ have \kl{positive coefficients}.
    In particular, $\Diff{K}(Y - K)$  belongs to $\CorrectPoly$.
    Furthermore, $\MaximalMonomials(\Diff{K}) = \MaximalMonomials(\Diff{K}(Y - K))
    \hoarele \MaximalMonomials(P)$.
    By induction hypothesis,
    we conclude that
    $\Diff{K}(Y - K)$ is \kl{represented}
    by a \kl{star-free $\Nat$-polyregular function} $g_1$,
    \begin{equation*}
        f_1(Y, \vec{X}) \defined
        \begin{cases}
            0 & \text{ if } Y \leq K \\
            \Diff{K} & \text{ otherwise }
        \end{cases}
        \quad .
    \end{equation*}

>>>>

When multiple indeterminate are involved, a bit of care is needed. The key
ingredient is to notice that polynomials in $\CorrectPoly$ are actually stable
under a \emph{discrete} analogue of the usual differentiation operator.
Intuitively, polynomials $P$ in $\CorrectPoly$ have \kl{positive} \kl{maximal
monomials}, and so does $\partial P$, $\partial^2 P$ etc.

\begin{definition}
    \label{discrete-derivative:def}
    Let $\vec{X}$ be a tuple of indeterminate,
    $K \in \Nat$,
    and 
    $P \in \Rel[\vec{X}]$ be a polynomial.
    \begin{equation*}
        \Diff{K}{P} \defined 
        P(\vec{X}) - P(\vec{X} - K) \quad .
    \end{equation*}
    Where $P(\vec{X} - K)$ is the polynomial function
    obtained by pre-composing $P$ with the map 
    $\vec{X} \mapsto \seqof{X_i - K}{X_i \in \vec{X}}$.
\end{definition}

As one would expect, the discrete derivatives are linear operations on
polynomials, that commutes with the partial application operators
$\restr{\cdot}{\nu}$.

\begin{fact}
    \label{discrete-deriv-linear:fact}
    For all tuple $\vec{X}$ of indeterminate,
    for all $K \in \Nat$, for all partial function
    $\nu \colon \vec{X} \topartial \Nat$:
    \begin{enumerate}
        \item $\Diff{K}{ \cdot } \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$
            is a linear operator,
        \item $\restr{\cdot}{\nu} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$
            is a linear operator,
        \item $\Diff{K}{ \cdot } \circ \restr{\cdot}{\nu}
            = \restr{\cdot}{\nu} \circ \Diff{K}{\cdot}$.
    \end{enumerate}
\end{fact}

The upcoming \cref{lem:correct-to-n-rat} is proven using an induction on the
number $d$ of indeterminate, and the degrees of the \kl{maximal monomials}. To
make the induction formal, let us define $\Monomials[\vec{X}]$ as the
(infinite) set of all possible \kl{monomials} over $\vec{X}$, ordered by
\kl{divisibility}. Remark that this set is isomorphic to $\Nat^{\vec{X}}$ with
the product ordering, which is a \kl{well-quasi-ordering}.

@@

Let us now prove that the \kl{discrete differentiation} operator
can be used to perform an induction.

\begin{fact}
    \label{derivation-simplifies:fact}
    For all $P \in \Rel[\vec{X}]$ that are non-constant,
    for all $K \in \Nat$,
    $\MaximalMonomials(\Diff{K}{P}) \hoarele
    \MaximalMonomials(P)$.
\end{fact}

\begin{fact}
    Let $M,T$ be two \kl{monomials}, such that
    $T$ \kl{strictly divides} $M$.
    Then,
    $T$ \kl{divides} some \kl{maximal monomial}
    of $\Diff{K}{M}$.
\end{fact}

\begin{lemma}
    \label{derivation-stabilises-correct:lem}
    Let $P \in \CorrectPoly$,
    $P_1$ be the sum of \kl{maximal monomials} of $P$,
    and $P_2 \defined P - P_1$ be the sum of
    non-maximal monomials of $P$.
    There exists a computable $K$,
    such that
    $(\Diff{K}{P_1} + P_2) \in \CorrectPoly$.
\end{lemma}

@@

    \textbf{Induction:} Assume that $P$ is not a constant polynomial.
    Leveraging \cref{derivation-stabilises-correct:lem}, there exists a
    computable $K \in \Nat$ such that $\Diff{K}{P} \in \CorrectPoly$. Using
    \cref{derivation-simplifies:fact} (because $P$ is non-constant), we also
    know that $\MaximalMonomials(\Diff{K}{P}) \hoarele \MaximalMonomials(P)$.
    By induction hypothesis, we conclude that $\Diff{K}{P}$ is \kl{represented}
    by a \kl{star-free $\Nat$-polyregular function} $f_1$.

    Assume that $\alpha Y^p M \in \MaximalMonomials(P)$
    for some indeterminate $Y$, $\alpha \neq 0$, and $p \geq 1$.
    Because $P \in \CorrectPoly$, we know that $\alpha > 0$.
    Using \cref{lem:delta-cool},
    we know that there exists a computable $K$ such that
    the polynomial
    $\Diff{K}{P} \defined P - \alpha (Y - K)^p M$
    satisfies that for all partial valuations $\nu \colon Y\vec{X} \topartial \Nat$
    such that $\nu(Y) > K$ or $\nu(Y)$ is undefined,
    the \kl{maximal monomials} of $\Diff{K}$ have \kl{positive coefficients}.
    In particular, $\Diff{K}{P} (Y - K)$  belongs to $\CorrectPoly$.
    Furthermore, $\MaximalMonomials(\Diff{K}{P}) = \MaximalMonomials(\Diff{K}{Y - K})
    \hoarele \MaximalMonomials(P)$.
    By induction hypothesis,
    we conclude that
    $\Diff{K}{Y - K}$ is \kl{represented}
    by a \kl{star-free $\Nat$-polyregular function} $g_1$,
    \begin{equation*}
        f_1(Y, \vec{X}) \defined
        \begin{cases}
            0 & \text{ if } Y \leq K \\
            \Diff{K} & \text{ otherwise }
        \end{cases}
        \quad .
    \end{equation*}
